Servants, Syndicates, and Saviors: The Rise of Domestic Political Interventionism by Senior Military Leaders (Prospectus)

### 2020-12-04 code/prospectusv2.Rmd
commit

---
title: "Servants or Syndicates?: The Rise of Domestic Political Interventionism by Senior Military Leaders (Prospectus)"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |

  _This dissertation explores changes in how the contemporary military institution intervenes in domestic politics. I advance a theory which argues that normative beliefs of military officers regarding political activism is a result of two interacting factors: the level of domestic political polarization within a society, and the degree of prestige and credibility the military holds. To support my argument, I trace changes in levels and types of distinct forms of political activism by military officers, including authoring opinion-editorials in major newspapers, as domestic polarization and military institutional credibility rises. I ultimately argue that the increasing political activism of retired military leaders is not merely a response to the hyper-partisanship present in the current political atmosphere, but rather the logical outworking of historical trends that have been building for decades. I conclude by describing the implications of changing beliefs among military leaders regarding the role the military should have in domestic politics. This dissertation contributes to the fields of international relations and civil-military relations by advanving an empirical and historical analysis of a growing phenomenon: military leaders intervening in domestic American politics. _

@@

Thankfully, within the United States, unlike many other countries across the world and throughout history, there has never been a coup, nor a serious attempt made by the military to dislodge political control from their duly-constituted elected leaders and to hold it themselves. However, as almost every civil-military relations scholar points out, the absence of coups does not mean that the relationship between civilian and military leaders is healthy, nor absent of conflict altogether. Indeed, a cursory examination of the relationships between Lincoln and MacClellan, Truman and MacArthur, and Obama and McChrystal reveals that disagreement between those who serve at the highest levels of civilian government and military organizations is neither rare nor understudied.

What has changed, however, is the degree to which a very specific subset of the American military institution - retired senior military officials - have participated in the domestic political process post retirement. Indeed, whereas General Omar Bradley, one of the military's five star generals and a hero of World War Two, famously remarked in 1959, "I am convinced that the best service a retired general can perform is to turn in his tongue along with his suit and mothball his opinions," [@times_bradley_1959], retired generals and admirals today appear frequently in the media, endorse Presidential candidates at higher frequency [@griffiths_not_2019], and frequently serve in government after their retirement from military service [@brooks_dont_2016]. To provide one such example, following protests across the United States in the summer of 2020 in which President Donald Trump nearly ordered the use of active duty military forces in American cities, former Chairman of the Joint Chiefs of Staff Admiral (Ret) Michael Mullen was one of at least ten four star retired officers who publicly aired concern about the President's potential use of force. ^[Other officers included General (Ret) James Mattis, General (Ret) Mark Kelly, Admiral (Ret) William McRaven, General (Ret) Martin Dempsey, General (Ret) Colin Powell, General (Ret) Michael Hayden, General (Ret) David Petraeus, General (Ret) John Allen, and General (Ret) Vincent Brooks.]. During an interview in the days which followed, when asked by NPR reporter why he spoke out, Mullen stated, "we were getting to a point where the country was in jeopardy, which is why I spoke out" [@isaacson_ret_2020; for further analysis, see also @brooks_let_2020].

Mullen's words are incredible from a number of important perspectives, and raise important normative questions regarding the military's role in a democratic society [@brooks_let_2020] and the conditions under which the military can and should express dissent to its civilian bosses [@cohn_what_2019]. Few, if any, scholars have attempted to systematically examine and explain the rise of political activism by retired military officers over time. Such an effort would better illumine the understanding of scholars, as well as the American military and society, of why a powerful group of what were once non-political actors have, at least in some cases, suddenly become one.  More important, a systematic study of this trend would contribute to a much-needed discussion regarding the role of the military institution and its relationship to the government and American society in a contentious political age.

The central argument made in this dissertation is that increasing political activism at the hands of military officers is the result of two longstanding and interconnected trends: rising political polarization within the domestic political sphere, and rising prestige and credibility of military institutions afforded by American society. The interaction of these trends explains the remarkable increase in political activity by retired senior military leaders, and portends a shift in the perceived role that these senior military leaders fill vis-a-vis civilian authority - a shift away from professional and obedient servants and towards "guardians of democracy" who pick and choose when and how to intervene in American politics. Such a shift builds on previous work regarding normative "role beliefs" [@fitch_armed_1998] of the military institution, and has significant ramifications on the organizational culture of the military, its relationship with government and society, and American national security.

@@

My proposed dissertation topic features within the subfield of civil-military relations, a broad academic subfield that is generally placed within the International Relations and Comparative Politics subfield (though within the American context, the subfield of American Politics also certainly applies). Civil-military relations involve a wide host of significant and important theoretical questions that involve the pull and push of power between institutions, the delineation of authority between military and civilian leaders, and the relationship between a state's military and society.

@@

Finer further classifies military intervention as one of four ascending levels. At the bottom or first level, which he calls "influence," Finer views the military as making appeals "to reason or emotion" in an attempt to persuade the civilian government to adopt a certain course of action [@finer_man_1962, 77]. Finer adds that intervention by the military at this level is not problematic and entirely "consistent with the supremacy of the civil power" [@finer_man_1962, 77]. The next level of intervention Finer labels "pressures or blackmail," describing efforts by the military to "convince the civil power by threat of some sanction" [@finer_man_1962, 77]. Finer adds that military intervention via this level is "on the margin," referring to his belief that while some cases may be unproblematic, many other forms, such as the "threat to stage a coup," are normatively wrong [@finer_man_1962, 77]. The third level of intervention, "displacement," refers to the "removal" of government officials by the military as a result of "violence or the threat of violence," and the fourth level, "supplantment," occurs when the military "sweeps away the civilian regime and establishes the military in its place" [@finer_man_1962, 78].

@@

But a more central concern of mine is understanding how and why political polarization might impact how the military views its central role in American politics. I argue that there are at least three ways through which increasing polarization can shape the apolitical and non-partisan ethos of the military, and thus, can alter how the military views its role in politics. The first is that there is increasing difficulty for military leaders to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of Huntingtonian military professionalism. Fewer and fewer issue areas and policy matters are outside of the "spell" of polarization, owing to a phenomenon schoalrs of American politis refer to as "conflict extension" [@layman_party_2002]. In previous eras of polarization, whereas a single issue may have served as the "battleground" for partisan conflict, this is no longer true today, as research has consistently shown that voters who identify as liberals are more liberal across multiple issue categories and that the same trend holds for conservatives [@abramowitz_great_2018]. As a result, whereas the prospect for cooperation on issues that were not politically polarized remained high in the past, today the prospect for cooperation in general is far lower [@layman_activists_2010].

@@

With respect to the military institution itself, one of the most massive and influential changes that has occurred over the past roughly seventy years is the increase in overall prestige, credibility, and influence of the military institution. While key legislative acts passed since World War Two, to include the National Security Act of 1947 and the Goldwater-Nichols Act of 1946, may have fixed important gaps in America's ability to meet its national security requirements [@inboden_national_2017], they also cemented the military as one of if not the dominant institution and actor within American politics. Over the same time period, while there have been spikes as a result of exogenous events such as wars and scandals that have involved the military, public opinion towards the military has risen sharpy. Consider that in the late 1970s, as a likely result of public opinion over the Vietnam War, 54% of Americans indicated that they trusted the military "a great deal or quite a lot," whereas in 2019, 73% of Americas expressed the same sentiments [@gallup_polls_confidence_2007]. Unsurprisingly given these developments, compared to other government institutions, today the American public trusts the military to a remarkably high degree. According to a recent Pew Research Center poll, 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018].

While military leaders are rightfully cognizant of the benefits of holding the public's trust, a number of scholars and practitioners have noted that the trend of "military worship" is not only unhealthy [@fallows_tragedy_2014], but likely harmful for democracy [@friend_military_2018]. Esteeming the military more than it ought can perpetuate, according to Mara Karlin and Alice Friend, a number of harmful myths, including the idea that military service is somehow superior than other, non-military forms of service [@friend_military_2018]. Over the long haul, a military that is perceived to be more prestigious than its normative civilian superiors can, according to numerous scholars, be problematic for civilian control of the miltiary [@friend_military_2018].

Scholars have also shown that when elites possess a high degree of credibility, their ability to influence and shape public opinion is likewise high. In his influential study, _The Nature and Origins of Mass Opinion_, John Zaller argues that formations in mass public opinion are the result of three interacting components: "variation in the information carried in elite discourse, individual differences in attention to this information, and individual differences in political values and other predispositions" [@zaller_nature_1992]. Zaller argues that throughout modern America, most citizens, as non-elites, must rely heavily on the messages that come from elites regarding various topics and issue areas [@zaller_nature_1992]. Zaller further argues that the messages that elites send is necessarily "stereotyped" in some way: that is, the messages that elites send, in order to be understood by regular citizens who are part of the public, must be "short, simple, and highly thematic" [@zaller_nature_1992]. In developing this argument, Zaller is not claiming that regular citizens are uninformed or careless people who are incapable of processing complex information, but rather acknowledging that most citizens care most about a world that is not "beyond their personal experience" [@zaller_nature_1992]. Thus, his main point is simply that elites play a significant role in forming and swaying public opinion.

Therefore, as the public's trust in the military has grown in the years since Vietnam, so has the military's potential to impact public opinion. This is especially true of military elites, who hold substantial positions of authority. An increase in credibility of the institution may not indicate that the military or particular military elites will actually engage in behavior to deliberately impact public opinion, but rather that the potential to do so has increased, as the American people tend to listen to those whom they trust and revere.

@@

The central question I wish to explore in my dissertation is whether and why the US military, or portions of it, have become increasingly politically active, and whether such a shift portends a shift in the military's own "role beliefs" concerning how it relates to American government and society. As such, my dissertation will center on fundamental and longstanding debates about the concepts of authority, loyalty, and the proper role of the military with respect to government. What is different from previous works in the field is that I wish to explore changes in these concepts, particularly within the US military, over time and in historical perspective.

While there have been several vital studies, most of them incorporating original survey data, exploring the beliefs that military members have regarding forms of political behavior [@dempsey_our_2010; @urben_party_2013; @urben_wearing_2014], these studies have not sought to explain these changes. Similarly, there have also been a number of recent and important studies regarding the impacts of political polarization on the credibility society places in the military as an instition [@robinson_michael_danger_2018; @robinson_who_2019], as well as a host of impressionistic pieces noting the potentially detrimental impacts of polarization and politicization on the military institution [@feaver_we_2016; @barno_how_2016; @golby_jim_keep_2020; @reid_retired_2020 ]. However, these studies also have not empirically explored the impact of increasing polarization and institutioncal prestige on the military as an institution over time. I wish to close that gap, as I think doing so will help not only scholars, but also practitioners, to include members of the military, understand the answers to some very important questions.  These questions include:

* Among whom within the military institution is a shift in normative "role beliefs" occurring, and why?
* When did such a shift start, and why? 
* If there is a shift in the military's normative "role beliefs," how will that impact the military's organizational culture?
* Are some of the most polarized domestic issues compatible with historically normative conceptions of military professionalism, such as the principles of civilian supremacy and a non-partisan ethos? why or why not?
* What are the key challenges or obstacles to the military regaining and sustaining an effective non-partisan ethos?

Specifically, I desire to explore the relationship between polarizing domestic politics and institutional credibility on the professional identity and character of the military. In the post World War Two era in particular, the military's institutional power, social prestige, and level of credibility has generally grown, making the military a recognized and respected actor within American politics. And, especially since the 1970s, political polarization has also increased dramatically, to the point that America is now as polarized as it was at the end of the American Civil-War [@hare_polarization_2014]. I believe that these two trends, occurring at least in a broad sense simultaneously, have significantly impacted the military's beliefs about its proper role in politics, and that a systematic investigation of this relationship should be undertaken.

---
### Figure this out

In the context of the US military, to my knowledge, loyalty has rarely been used as a conceptual construct in the political science literature, likely and thankfully because the US military has not yet considered, or been involved, in a coup. Nonetheless, I contend that there is a rich opportunity to study strategic loyalty as a concept, especially in the contemporary context of the United States' domestic political climate, as political polarization has increased extensively.

I argue that this third challenge is especially relevant for the contemporary military, and specifically, with respect to how members of the military exercise the concept of loyalty. As an increasing number of what individuals believe to be core issues are raised and debated in the partisan arena, individuals face decisions of how to prioritize the persons or objects of their loyalty and further, what actions to take on the basis of loyalty to these persons or objects. I further theorize (though this has to be tested, of course) that while all individuals will likely face some version of this challenge of loyalty in a polarizing age, the challenges will be more acute for military leaders. Not only is the de-facto setting in which military leaders operate marked by multiple and often conflicting loyalties due to the structure of and statutes set forth in the US Constitution, but also, the potential consequences are higher for military leaders than average citizens should they fail to discharge the duties that correspond to their roles as professional officers. These consequences include failure to achieve tangible national security goals, such as loss on the battlefield, or less tangible but still important outcomes such as maintaining trust with the American public.

As important as these consequences are, they do not speak to another set of consequences individual military leaders also may face in a polarizing age. Indeed, in addition to consequences to the country that may occur should the military fail to discharge its duties to safeguard America, military leaders also potential consequences if confronted with a situation in which they have to execute orders which are technically legal, but morally borderline at best. Military leaders are still "moral agents" who "remain responsible for the decisions and actions they take" [@dubik_just_2016, 160]. But the considerations of what constitutes moral right and wrong to begin with may be changing as polarization worsens [@graham_liberals_2009].

Therefore, I argue that if polarization continues to heighten the identities of groups that people associate with, as well as come closer to the boundaries of how individuals think about moral right and wrong, there is an increased likelihood of military leaders having to choose between conflicting loyalties. What happens if and when the objects of a military leader's multiple loyalties not merely disagree, but become stubbornly divided ideologically? Can such an outcome exist? Is there empirical proof that such a trend is happening? If so, what are the normative implications, both for the military as a profession, and for the nation's security, and are there actions that can be taken to halt such a trend?

 In the past several years for example, there has been a concerted effort by the services to identify and punish senior leaders who engage in misconduct, to demonstrate to Congress that senior military leaders do not need assistance to handle sexual assault, and to otherwise prove that it can be trusted to accomplish its missions [@us_army_repairing_2014]. Former Chief of Staff of the Army, Retired General Ray Odierno, told fellow Army generals that "of great importance within the Army Profession is the duty of each of us to be a steward of our profession and to maintain its essential and hard won trust with society [Ray Odierno in @us_army_repairing_2014].

---

 I am specifically interested in the question of if and how political polarization impacts the concept of loyalty in the military. This is an interesting research question because it brings together at least three relevant strands of literatures which, when combined, generates an interesting set of puzzles and questions. First, one of the most fundamental normative hallmarks of the military's professional nature, at least by Huntingtonian standards, is that it ought to be an apolitical and non-partisan institution. Second, one of the features of the US military is that it prizes and elevates the trait of loyalty among those who serve and, by statute as well as normative tradition, exercises simultaneous dual loyalties to its civilian heads, the President as the Commander in Chief, and Congress. And third, we know from the American politics literature that ideological polarization and partisan loyalty are connected in that those citizens who hold to stronger ideological convictions tend to be the most loyal partisans in terms of voting and other political behaviors [@abramowitz_great_2018, 112-116].

If these things are true, then a question is whether rising political polarization will generate instances in which military officers must choose between which objects of loyalty to ultimately elevate and side with in the event that there is conflict between and among these objects. Moreover, at the level of the military institution, a similar problem may also emerge: will the military institution have to choose between competing objects of loyalty, such as the President as the Commander in Chief, the Congress, and even its own professional ethic?^[Loyalty had to be a prime consideration in several prominent, recent cases involving the military and the government. Consider the case of now retired Army Lieutenant Colonel Alex Vindman, who testified before Congress against the US President, Donald Trump, regarding what Vindman believed to be behavior by the President that may have been illegal, or the case of Navy Captain Brett Crozier, who leaked a letter to the press after his pleas for help in battling the coronavirus aboard the ship he commanded, the USS Theodore Roosevelt, went unheeded for what he felt was too long. Certainly these instances involved individuals making choices about to whom or to what they would be most loyal to - the President, the Constitution, the ideals and obligations of the military, the operational mission, themselves as individual moral agents, and perhaps even more objects of loyalty.]

Other questions I wish to explore include whether there patterns in the way that loyalty obligations held by the military are in fact challenged by governmental actors during periods of increased political polarization, and if so, what are they, and how is the military reacting to these patterns? Are there impacts on the military's professional identity as a result of political polarization, and if so, from who or what? Are partisan/ideological loyalties displacing other loyalty relationships within the military?  Through what mechanisms do these changes operate? What sorts of changes are they driving, if any, in the military institution? Is this problematic for the military profession, and why? These and other questions will shed light on how rising political polarization is impacting the professional identity of the military.

The gaps in the current literature, which I hope to advance and fill, involve the impact of political polarization directly on the military institution itself, and why the military experiences such impacts. A number of excellent and recent scholarly works have looked at aspects of this question, but from an angle that examines not the military directly, but rather society's perception of the military or various aspects of the military as a result of polarized politics. A number of these works have directly examined the impact of partisan affiliation on the credibility of military actors, resulting in some very interesting findings. Michael Robinson, for instance, found that members of the public did not sanction military elites for making public messages that accorded with a member's partisan views, and that the public only sanctioned those elites whose messages were in line with the opposite political party [@robinson_danger_2018]. Previously, Heidi Urben found that among active military officers, there was weakening support for the norm that retired military officers should remain apolitical or non-partisan in retirement [@urben_party_2013]. These scholars' findings suggest, as others have pointed out [@brooks_paradoxes_2020], that political polarization at least coincides with, if not causes, a weakening in support among both the public and among service members for the military's normative apolitical ethos. Other scholars have also unpacked a trove of information regarding public opinion and the military, including how partisan affiliation shapes differences in support for the use of military force [@golby_jim_elite_2017] and military elite endorsement of political candidates [@golby_military_2012].

Focusing attention on the military institution will help scholars in the field of civil-military relations and international relations more broadly better understand if and how the military, as a unique body charged with the defense of the nation, changes in fundamental ways as a result of a polarized political climate. On one hand, one would expect that it might, in the sense that the American military is, after all, a microcosm of American society. Therefore, one could expect that if society truly is becoming polarized, then at least to some degree, so is the military, or at least those who volunteer for service and at the point in time that they join. On the other hand, one might reasonably expect that given the unique purpose of the military, combined with its long-standing professional ethic, might insulate the military institution from some of the potentially negative impacts of polarized politics. These differing expectations need to be more clearly articulated and explored.

To be clear, scholars have long studied how domestic politics impact the military. And as scholars have pointed out, military and civilian relationships are in tension by their very nature in that the leaders of a democracy allow a most undemocratic institution, the military, to protect the polity [@feaver_civil-military_1996]. And as American history proves, there have been countless instances in which civilian and military leaders disagreed. Many of these rise to the level of common knowledge: Lincoln fired McClellan because he lost confidence in him as a wartime Commander; Truman sacked MacArthur because the latter was insubordinate regarding the future of war in Asia; and Obama and McChrystal clashed about troop levels in Afghanistan, ultimately resulting in the former firing the latter. Beyond these episodes in which President and wartime commander locked horns, consider several other instances of "civilians" and "military" clashing. One episode is certainly the way in which the Vietnam war escalated in ways that, at least according to H.R. McMaster, military generals at the time knew were destined to fail [@mcmaster_dereliction_1998]. A more recent example, and one much studied and commented upon by civil-military relations scholars, is the so-called "Revolt of the Generals" that occurred following the early years of the invasion of Iraq, which involved several recently-retired generals and admirals sharply criticizing Secretary of Defense Rumsfeld's leadership dynamics [@cook_revolt_2008].

My intended research topic is different than these other, still very important topics in which disagreements between the executive and the military occur during wartime. It is my assessment that polarization is causing conflict, or certainly has the potential to do so, between the military and civilian leaders, but in most instances, the subject is about topics other than the use of military force, which is somewhat puzzling, given the military's reason d'etre. For example, I believe we are seeing portions of the military institution (particular retired military officers) openly disagree with civilian leaders on topics that I believe are, even by the widest of academic definitions, outside the bounds of what constitutes appropriate military expertise.^[I hope to prove this, empirically, in or as part of a chapter in my dissertation]

This raises, then, an important question: does the military's essential purpose risk evolving or changing significantly during polarized times? While I believe that these potential developments are catching the attention of civil-military relations scholars, none to my knowledge have specifically set out to systematically explore if and how polarized partisan politics are impacting the military's professional identity and character, and in particular, the role of the military institution itself in facilitating these developments. Most scholars acknowledge the potentially damaging impacts of the politicization of the military [@golby_listening_2013; @brooks_paradoxes_2020] at the hands of civilian leaders, and I share these concerns. Fewer scholars, if any, are exploring the potential role of the military in politicizing itself, and the mechanisms through which these developments occur.

## Military Professions and Organizational Loyalty

As a subject or concept, loyalty features prominently in the international relations and civil-military relations literatures, often in a comparative setting and in the context of coups [@marcum_overthrowing_2016; @johnson_squeaky_2018], civil wars [@mclauchlin_loyalty_2018], and the difficulties that multi-ethnic states face in maintaining a professional military [@peled_question_1998]. This literature deals with a host of important issues that stem from problems associated with ethnic identification and loyalty, either to a regime, cause, or leader. Though the US context is very different from the contexts usually explored in this literature, its insights may help me identify related dynamics and/or mechanisms that are unfolding in the US: how are changes in partisan or ideological identification (instead of ethnic) as a result of political polarization impacting conceptions of loyalty within the US military, if at all? To even begin answering this question, one has to first define what loyalty is and how it works. For this fundamental task, the fields of psychology, ethics, management, and sociology are helpful.

Loyalty has been conceptualized as both a virtue [@coleman_problems_2009] and as a moral emotion [@connor_military_2019], the main difference between the two being that the former is a normative aspirational trait, while the latter is a primed behavioral response. For the purposes of my dissertation, these differences are not consequential, because both assert that loyalty is something that is ultimately felt by an individual and subsequently reflected in action, such as, in the military context, "the desire to protect each other from harm and to defend each other from attack" [@coleman_problems_2009, 109]. Oldenquist, a philosopher, further describes loyalty as a devotion or attachment that an individual has to someone or something specific such that one refers to the object of loyalty as "mine" and is inclined to feel "pride when it prospers, shame when it declines, and anger or indignation when it is harmed" [@oldenquist_loyalties_1982, 175]. Loyalty also generates a willingness in the individual to assume "obligations" [@oldenquist_loyalties_1982, p. 175] on behalf of the object of one's loyalty, and these obligations can take multiple forms, such as a willingness of the individual to endure hardships and make sacrifices [@fielder_organizational_1992]. Scholars agree that individuals can hold multiple loyalties simultaneously, and recognize that this often can and does cause problems [@coleman_problems_2009]. For organizations, and by extension, professions, the concept of loyalty is also very important. Conner et al. state, for example, that "having loyal members is vital to perpetuating the group" [@connor_military_2019].

>>>>

---
title: "Servants, Syndicates, and Saviors: The Rise of Domestic Political Interventionism by Senior Military Leaders (Prospectus)"
author: "Peter M. Erickson"
date: |
  `r format(Sys.time(), '%B %d, %Y')`
abstract: |

  _This dissertation explores changes in how the contemporary military institution intervenes in domestic politics. I advance a theory which argues that normative beliefs of military officers regarding political activism is a result of two interacting factors: the level of domestic political polarization within a society, and the degree of prestige and credibility the military holds. To support my argument, I trace changes in levels and types of distinct forms of political activism by military officers, including authoring opinion-editorials in major newspapers, as domestic polarization and military institutional credibility rises. I ultimately argue that the increasing political activism of military leaders is not merely a response to the hyper-partisanship present in the current political atmosphere, but rather the logical outworking of trends involving the prestige of the military institution as well as political polarization that have been building for decades. I conclude by describing the implications of changing beliefs among military leaders regarding the role the military should have in domestic politics. This dissertation contributes to the fields of international relations and civil-military relations by advancing an empirical and historical analysis of a growing phenomenon: military leaders intervening in domestic American politics. _

@@

Thankfully, within the United States, unlike many other countries across the world and throughout history, there has never been a coup, nor a serious attempt made by the military to dislodge political control from their duly-constituted elected leaders and to hold it themselves. However, as almost every civil-military relations scholar points out, the absence of coups does not mean that the relationship between civilian and military leaders is healthy, nor absent of conflict altogether. Indeed, a cursory examination of the relationships between Lincoln and McClellan, Truman and MacArthur, and Obama and McChrystal reveals that disagreement between those who serve at the highest levels of civilian government and military organizations is neither rare nor understudied.

What has changed, however, is the degree to which a very specific subset of the American military institution - senior military leaders - have participated in the domestic political process, especially post retirement. Indeed, whereas General Omar Bradley, one of the military's five star generals and a hero of World War Two, famously remarked in 1959, "I am convinced that the best service a retired general can perform is to turn in his tongue along with his suit and mothball his opinions," [@times_bradley_1959], retired generals and admirals today appear frequently in the media, endorse Presidential candidates at higher frequency [@griffiths_not_2019], and frequently serve in government after their retirement from military service [@brooks_dont_2016]. To provide only one such example, following protests across the United States in the summer of 2020 in which President Donald Trump nearly ordered the intervention of active duty military forces into American cities, former Chairman of the Joint Chiefs of Staff, Admiral (Ret) Michael Mullen, was one of at least ten four star retired officers who publicly aired concerns through a variety of mediums about the President's potential use of active-duty military force. ^[Other officers included General (Ret) James Mattis, General (Ret) Mark Kelly, Admiral (Ret) William McRaven, General (Ret) Martin Dempsey, General (Ret) Colin Powell, General (Ret) Michael Hayden, General (Ret) David Petraeus, General (Ret) John Allen, and General (Ret) Vincent Brooks.]. In fact, during an interview in the days which followed the incident, when asked by an NPR reporter why he decided to publicly share his thoughts on the situation, Mullen stated, "we were getting to a point where the country was in jeopardy, which is why I spoke out" [@isaacson_ret_2020; for further analysis, see also @brooks_let_2020].

Mullen's words are incredible from a number of perspectives, and raise important normative questions regarding the military's role in a democratic society [@brooks_let_2020] and the conditions under which the military can and should express dissent to its civilian bosses [@cohn_what_2019]. And while a small but dedicated number of civil-military relations scholars do examine these and related issues quite frequently in today's charged political environment, few, if any, have attempted to systematically examine and explain over time the conditions under which senior military leaders have broken and will break normative traditions and engage in domestic politics. Such an effort would better illumine the understanding of scholars, as well as of the military and society more broadly, concerning why a powerful group of what were once non-political actors have, at least in some cases, suddenly become one.

The central argument made in this dissertation is that increasing political activism at the hands of military officers indicates a shift in the beliefs of the military regarding its central role and purpose in society, or "role beliefs" [@fitch_armed_1998]. This shift, moreover, is the result of two longstanding and interconnected trends: rising political polarization within the domestic political sphere, and the rising prestige and credibility of military institutions afforded by American society. The interaction of these trends explains the remarkable increase in political activity by retired senior military leaders, and portends a shift in the perceived role that these senior military leaders fill vis-a-vis civilian authority - a shift away from professional and obedient "servants" and towards that of "syndicates" or even "saviors" of the American state. This shift, moreover, holds enormous ramifications regarding the organizational culture of the military, its relationship with government and society, and American national security.

@@

My proposed dissertation topic features within the sub-field of civil-military relations, a broad academic sub-field that is generally placed within the International Relations and Comparative Politics sub-field (though within the American context, the sub-field of American Politics is entirely relevant and appropriate). Civil-military relations involve a wide host of significant and important theoretical questions that involve the pull and push of power between institutions, the delineation of authority between military and civilian leaders, and the relationship between a state's military and society.

@@

Finer further classifies military intervention as one of four ascending levels. At the bottom or first level, which he calls "influence," Finer views the military as making appeals "to reason or emotion" in an attempt to persuade the civilian government to adopt a certain course of action [@finer_man_1962, 77]. Finer adds that intervention by the military at this level is not problematic and entirely "consistent with the supremacy of the civil power" [@finer_man_1962, 77]. The next level of intervention Finer labels "pressures or blackmail," describing efforts by the military to "convince the civil power by threat of some sanction" [@finer_man_1962, 77]. Finer adds that military intervention via this level is "on the margin," referring to his belief that while some cases may not be problematic, many other forms, such as the "threat to stage a coup," are normatively wrong [@finer_man_1962, 77]. The third level of intervention, "displacement," refers to the "removal" of government officials by the military as a result of "violence or the threat of violence," and the fourth level, "supplantment," occurs when the military "sweeps away the civilian regime and establishes the military in its place" [@finer_man_1962, 78].

@@

But a more central concern of mine is understanding how and why political polarization might impact how the military views its central role in American politics. I argue that there are at least three ways through which increasing polarization can shape the apolitical and non-partisan ethos of the military, and thus, can alter how the military views its role in politics. The first is that there is increasing difficulty for military leaders to remain politically neutral, non-partisan, or otherwise apolitical - a hallmark of Huntingtonian military professionalism. Fewer and fewer issue areas and policy matters are outside of the "spell" of polarization, owing to a phenomenon scholars of American politics refer to as "conflict extension" [@layman_party_2002]. In previous eras of polarization, whereas a single issue may have served as the "battleground" for partisan conflict, this is no longer true today, as research has consistently shown that voters who identify as liberals are more liberal across multiple issue categories and that the same trend holds for conservatives [@abramowitz_great_2018]. As a result, whereas the prospect for cooperation on issues that were not politically polarized remained high in the past, today the prospect for cooperation in general is far lower [@layman_activists_2010].

@@

With respect to the military institution itself, one of the most massive and influential changes that has occurred over the past roughly seventy years is the increase in overall prestige, influence, and public credibility of the military institution. While key legislative acts passed since World War Two, to include the National Security Act of 1947 and the Goldwater-Nichols Act of 1946, may have fixed important gaps in America's ability to meet its national security requirements [@inboden_national_2017], they also cemented the military as one of the dominant institutions and actors within American politics. Over the same time period, while there have been spikes as a result of exogenous events such as wars and scandals that have involved the military, public opinion towards the military has increased. Consider that in the late 1970s, as a likely result of public opinion over the Vietnam War, 54% of Americans indicated that they trusted the military "a great deal or quite a lot," whereas in 2019, 73% of Americas expressed the same sentiments [@gallup_polls_confidence_2007]. Compared to other government institutions, today the American public trusts the military to a remarkably high degree. According to a recent Pew Research Center poll, 80% of the American public expects the military to act in the best interests of the American public, compared to business leaders (45%), the media (40%), and elected officials (25%) [@johnson_trust_2018].

Coinciding with these developments, scholars began exploring whether there was an unhealthy "gap" forming between American society and the military, especially after the US ended conscription and formed the all-volunteer force in 1973 [@feaver_soldiers_2001; @schake_warriors_2016]. While there is no definitive answer as to whether a broad "gap" exists, scholars have noted the existence of several smaller "gaps," including a "partisan gap" [@golby_duty_2011] between an officer corps that strongly (though not exclusively) identifies with the Republican party and potential Democratic presidential administrations, and a "familiarity gap" between an all-volunteer military that has endured repeated deployments since September 11, 2001 and American society, especially younger Americans [@golby_thanks_2016; @colford_young_2016]. A major concern of scholars is that to the degree that these gaps exist, these potentially facilitate the formation of unhealthy attitudes regarding the superiority of military service over and above other types of public service, and further reinforces the military's isolation from society [@fallows_tragedy_2014] while also negatively impacting democracy [@friend_military_2018]. According to Mara Karlin and Alice Friend, the act of "military worship" - esteeming the military more than society ought to - can, over the long haul, be problematic for civilian control of the military [@friend_military_2018].

We further know, as scholars have argued, that when elites possess a high degree of credibility, they possess the ability to influence and shape public opinion. In his influential study, _The Nature and Origins of Mass Opinion_, John Zaller argues that formations in mass public opinion are the result of three interacting components: "variation in the information carried in elite discourse, individual differences in attention to this information, and individual differences in political values and other predispositions" [@zaller_nature_1992]. Zaller argues that throughout modern America, most citizens, as non-elites, must rely heavily on the messages that come from elites regarding various topics and issue areas [@zaller_nature_1992]. Zaller further argues that the messages that elites send is necessarily "stereotyped" in some way: that is, the messages that elites send, in order to be understood by regular citizens who are part of the public, must be "short, simple, and highly thematic" [@zaller_nature_1992]. In developing this argument, Zaller is not claiming that regular citizens are uninformed or careless people who are incapable of processing complex information, but rather acknowledging that most citizens care most about a world that is not "beyond their personal experience" [@zaller_nature_1992]. Thus, his main point is simply that elites play a significant role in forming and swaying public opinion.

Therefore, as the public's trust in the military has grown, as gaps between segments of society and the all-volunteer force have emerged, and as military service has come to be viewed as somehow superior to other forms of civilian service, the latent power and influence of the military has grown significantly. This is especially true of military elites, who hold substantial positions of authority and responsibility. However, an increase in the power and influence of the military institution may not indicate that the military will actually use it in particularly harmful ways, but merely that it has the potential to do so, as the American people tend to listen to those whom they trust and revere.

@@

The central question I wish to explore in my dissertation is whether and why the US military, or portions of it, have become increasingly politically active, and whether such a shift portends a shift in the military's own "role beliefs" concerning how it should and does relate to American government and society. As such, my dissertation will center on fundamental and longstanding debates about the proper role of the military with respect to government. I wish to explore changes in these concepts over time and in historical perspective within the United States.

While there have been several rich and important studies that explore the attitudes of military officers towards various forms of political behavior [@dempsey_our_2010; @urben_party_2013; @urben_wearing_2014], these studies have not primarily sought to explain why these attitudes have changed over time. Similarly, while there have been a number of important studies exploring the impacts of political polarization on the credibility society places in the military as an institution [@robinson_michael_danger_2018; @robinson_who_2019], the dependent variable in these studies was the credibility that society places in the military. There have been a number of impressionistic pieces noting the potentially detrimental impacts of polarization and politicization on the military institution [@feaver_we_2016; @barno_how_2016; @golby_jim_keep_2020; @reid_retired_2020], but these works, too, have not sought to empirically explain the impacts of these changes over time. It is my opinion that civil-military relations scholars need grounding in theory, specific to the context of the United States, that seeks to explore how rising political polarization and military influence interact, and that such a theory needs to be supported by rigorous empirical work. I wish to close this gap, as doing so will enable scholars and practitioners, to include members of the military, to understand the answers to some very important questions, including:

* Has the answer to the question, "what does it mean to be a military professional?" changed over the past several decades, and if so, why?
* Among whom within the military institution is a shift in normative "role beliefs" occurring, and why?
* When did such a shift start, and have there been similar periods in American history?
* If there is a shift among portions of the military in "role beliefs," how does such a shift impact the military's organizational culture?
* Are some of the most contemporary polarizing domestic issues compatible with historically normative conceptions of military professionalism, such as the principles of civilian supremacy and a non-partisan ethos? Why or why not?
* What are the key challenges or obstacles to the military regaining and sustaining an effective non-partisan ethos? 
* Should we expect these challenges and obstacles to endure?

Answers to these questions will be of particular importance to furthering our understanding of how the US military has responded to epochs of polarizing politics before, and how changes in levels of institutional prestige impact these responses. I suspect that my dissertation will show that any contemporary shift in military role beliefs is not the first shift that has occurred in American history, that such a shift has long been underway, and that particular aspects of today's polarized climate portend that the military will wrestle with the issue of shifting role beliefs for years to come. For these reasons, and more, a systematic investigation of the impact of polarizing politics and institutional prestige on the US military's beliefs concerning its role in government should be undertaken.
